---
title: "Analysis of Mental Health Risk"
author: "Kayla Lu"
date: "2026-03-07"
output: html_document
---

# Introduction
Mental health is influenced by multiple demographic, lifestyle,and work-related factors.

This project explores the distribution of mental health risk and examines how different factors such as gender, sleep,
stress level, and work environment are associated with mental health outcomes.

# Data Preparation

```{r setup, message=FALSE, warning=FALSE}
# Load required packages
library(tidyverse)
library(ggplot2)

# Read dataset
mental_data <- read_csv("mental_health_dataset.csv")
```


## Data Inspection
```{r}
head(mental_data)
str(mental_data)
summary(mental_data)
colSums(is.na(mental_data))
```

## Data Cleaning
```{r}
mental_data$gender <- as.factor(mental_data$gender)
mental_data$employment_status <- as.factor(mental_data$employment_status)
mental_data$work_environment <- as.factor(mental_data$work_environment)
mental_data$mental_health_history <- as.factor(mental_data$mental_health_history)
mental_data$seeks_treatment <- as.factor(mental_data$seeks_treatment)

mental_data$mental_health_risk <- factor(
  mental_data$mental_health_risk,
  levels = c("Low","Medium","High")
)
```

# Data Visualisations and Analysis
## Sample Overview
### Age Distribution
```{r}
ggplot(mental_data, aes(x = age)) +
  geom_histogram(binwidth = 5, fill="skyblue", color="white") +
  labs (title="Age Distribusion",
        x= "Age", y = "Count") +
  theme_minimal()
```

The age distribution shows that most participants fall within the young to middle-aged adult range, suggesting that the dataset primarily represents working-age individuals.

### Gender Distribution
```{r}
ggplot(mental_data, aes(x=gender)) +
         geom_bar(fill="skyblue") +
         theme_minimal() +
         labs(title="Gender")
```

The dataset includes participants from different gender groups, with gender groups fairly evenly represented

## Core Analysis
### Define custom colors 
```{r}
risk_colors <- c(
  "Low" = "lightgreen",
  "Medium" = "khaki",
  "High" = "lightcoral"
)
scale_fill_manual(values = risk_colors)
```

### 1. Mental health Risk Distribution
```{r}
ggplot(mental_data, aes(x = mental_health_risk, fill= mental_health_risk)) + 
  geom_bar() +
  scale_fill_manual(values = risk_colors) +   
  geom_text(stat = "count",
    aes(label = ..count..),
    position = position_stack(vjust = 0.5)) +
  labs(title = "Distribution of Mental Health Risk",
    x = "Mental helath Risk", y = "Count") +
  theme_minimal() +
  theme(legend.position = "none") 
```

The distribution of mental health risk varies across participants, with most individuals falling into the **Medium** risk category.

### 2. Gender vs Mental Health Risk
```{r}
Gendaer_mental_porportion <- mental_data %>%
  count(mental_health_risk, gender) %>%
  group_by(gender) %>%
  mutate(prop = n / sum(n))

ggplot(Gendaer_mental_porportion, aes(x = gender, y = prop, fill = mental_health_risk)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = risk_colors) +  
  labs(title = "Gender vs Mental Health Risk",
       x = "Gender", y="Percent") +
  theme_minimal()
```

Mental health risk is similarly distributed across genders, with most participants at medium risk, then high, and the fewest at low risk.

### Mental Health Measures
### 3. Depression vs Mental Health Risk
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y = depression_score, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +  
  labs(title = "Depresion score vs Mental Helath Risk",
       x = "Mental Health Risk", y = "Depresion Score") +
  theme_minimal() +
  theme(legend.position = "none") 
```

Depression scores are clearly associated with mental health risk: high scores correspond to high risk, medium scores to medium risk, and low scores to low risk.

### 4. Anxiety vs Mental Health Risk
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y= anxiety_score, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +  
  labs(title = "Anxiety score vs Mental Health Risk",
       x = "Mental Health Risk", y ="Anxiety Score") +
  theme_minimal() +
  theme(legend.position = "none") 
```

Anxiety scores are also associated with mental health risk, with higher scores indicating greater risk, moderate scores corresponding to medium risk, and lower scores to low risk.

### 5. Seek treatment vs Mental health Risk
```{r}
Seek_treatment_mental_porportion <- mental_data %>%
  count(seeks_treatment, mental_health_risk) %>%
  group_by(mental_health_risk) %>%
  mutate(prop = n / sum(n))

ggplot(Seek_treatment_mental_porportion, aes(x = mental_health_risk, y = prop, fill = seeks_treatment)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("No" = "lightgreen", "Yes" = "lightcoral")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percantage of Seeking Treatment by Mental Health Risk",
       x = "Mental Health History", y = "Percentage", fill = "Seek Treatment") +
  theme_minimal()
```

Across all three mental health risk levels, the majority of participants have not sought treatment; however, the high-risk group has a slightly higher proportion of participants who have sought treatment compared to the other two groups.

### 6. Mental Health History vs Mental health Risk
```{r}
History_proportion <- mental_data %>%
  count(mental_health_history, mental_health_risk) %>%
  group_by(mental_health_risk) %>%
  mutate (prop = n / sum(n))
  
ggplot(History_proportion, aes(x = mental_health_risk, y = prop, fill = mental_health_history)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_fill_manual(values = c("No" = "lightgreen", "Yes" = "lightcoral")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percentage of Mental Health History by Mental health Risk",
       x  ="Mental Helath Risk", y = "Percentage" , fill = "Mental Health History") +
  theme_minimal()
```

The distribution of mental health history is similar across all three risk levels, with about two-thirds of participants reporting no history. The high-risk group has a slightly higher proportion of participants with no history compared to the other two groups.

### Health and Lifestyle Factors##
### 7. Sleep vs Mental Health Risk
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y = sleep_hours, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +
  labs(title = "Sleep hours vs Mental Health Risk",
       x = "Mental Health Risk", y = "Sleep Hours") +
  theme_minimal() +
  theme(legend.position = "none")
```

Sleep duration is similarly distributed across all three mental health risk levels, with most participants sleeping between 5.5 and 7.5 hours.

### 8. Stress Level
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y = stress_level, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +  
  labs(title = "Stress Level vs Mental Health Risk",
       x = "Mental Health Risk", y = "Stress Level") +
  theme_minimal() +
  theme(legend.position = "none")
```

The overall range of stress levels is consistent across all three risk levels, with median scores being notably higher for medium and high-risk participants.

### 9. Social Support vs Mental Health Risk
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y = social_support_score, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +
  labs(title = "Social Support Score vs Mental Health Risk",
       x = "Mental Health Risk", y ="Social Support Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

Social support scores are similar across all mental health risk levels.

### Work-Related Factors##
### 10. Employment Status vs Menatl Health Risk
```{r}
Employment_mental_propotion <- mental_data %>%
  count(employment_status, mental_health_risk) %>%
  group_by(mental_health_risk) %>%
  mutate(prop = n / sum(n))

ggplot(Employment_mental_propotion, aes(x = mental_health_risk, y= prop, fill = employment_status)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prop)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percentage of Emploment Status by Mental Health Risk",
       x = "Mental Health Risk", y = "Percentage") +
  theme_minimal() 
```

Employment status is generally similar across all mental health risk levels, with most participants being employed, followed by students, and the fewest unemployed or self-employed. In the high-risk group, the proportion of employed and students is slightly higher than in the other two groups.


### 11. Work Environment vs Mental Health Risk
```{r}
Environment_mental_proportion <- mental_data %>%
  count(work_environment, mental_health_risk) %>%
  group_by(mental_health_risk) %>%
  mutate(prop = n / sum(n))

ggplot(Environment_mental_proportion, aes(x = mental_health_risk, y = prop, fill= work_environment))+
  geom_col() +
  geom_text(aes(label = scales::percent(prop)),
            position = position_stack(vjust = 0.5), size = 3) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percentage of Work Environment by Mental Health Risk",
       x = "Mental Health Risk", y = "Percentage") +
  theme_minimal()
```

Overall, most participants work onsite, accounting for nearly half, followed by remote and hybrid work. However, in the high-risk group, the proportion of onsite workers is lower than in the other two groups.

### 12. Productivity vs Mental Health Risk
```{r}
ggplot(mental_data, aes(x = mental_health_risk, y = productivity_score, fill = mental_health_risk)) +
  geom_boxplot() +
  scale_fill_manual(values = risk_colors) +
  labs(title = "Productivity Score vs Mental Health Risk",
       x = "Mental Health Risk", y ="Productivity Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

The productivity scores show a distinct downward trend across the three risk levels. The median score drops noticeably from low to high risk, and the middle 50% of scores show almost no overlap between risk levels.


# Conclusion
This analysis explored demographic, lifestyle, and work-related factors associated with mental health risk.

The results suggest that higher **depression scores**, **anxiety scores**, and **stress levels** are associated with increased mental health risk. Additionally, there is a negative relationship between **mental health risk** and **productivity scores**, indicating that participants with higher mental health risk tend to have lower productivity.
This analysis explored demographic and lifestyle factors associated with mental health risk.

Overall, these findings highlight that **psychological wellbeing is the main driver of mental health risk**, while productivity is inversely related. Addressing psychological factors may help improve both mental health outcomes and work performance.


